 

### 2020-12-24 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

Light conditions, specifically photosynthetically active radiation (PAR) varies along the vertical gradient with canopy height, canopy structure, across species and forests [@koike_leaf_2001] due to cumulative layers of vegetation. Generally, upper canopies are more exposed to irradiance and limit light through shading to canopy interiors, lower canopy layers and understories [@fauset_differences_2018; @niinemetsPackingPhotosyntheticMachinery2009; Parker 1995; @poorterMetaanalysisPlantResponses2019b]
(Fig. `r fig_NEON_vertical`).
In heterogeneous canopies [with high gap fractions], distance from the outer canopy is a better proxy for light environment than height (Parker 1995) and lower canopy light environments are highly variable, ranging from high light gap to deeply shaded understory regions [@chazdonPhotosyntheticLightEnvironments1984; @tymenQuantifyingMicroenvironmentalVariation2017; @smithSeasonalDroughtrelatedChanges2019]. 
 Tropical forests with dense overstory limit light to understories and subsequent layers, where tropical forest floors receive approximately 1-2%  of the light incident to the top of the canopy [@robertsStomatalBoundaryLayerConductances1990a; @chazdonPhotosyntheticLightEnvironments1984; @parkerConsequencesEnvironmentalHeterogeneity2019], whereas mixed-wood boreal and Northern temperate forest floors receive around 4-5% of full sunlight [Bartemucci et al. 2005].
This difference in light regime along a vertical profile is an important driver of plasticity in leaf traits, leaf physiology, adaptation, and thermal sensitivity along the gradient [@michaletzEnergeticCarbonEconomic2016; @fausetDifferencesLeafThermoregulation2018; @niinemetsResponsesForestTrees2010, @sackHowStrongIntracanopy2006; @doughtyAreTropicalForests2008].

@@

Diurnal moderating capacity of under forest microclimate differs between forests with varying canopy structure, crown density, and seasons. Tropical, and temperate broadleaved and non-pine conifer forests maintain lower daytime maximum temperatures than pine and boreal forests. Increase in crown density and leaf area index increases decreases light and higher maximum temperature permeability in forests, thus maintaining lower maximum daytime temperature. Decrease in crown density and leaf area index such as in pine forests that have sparse stands decreases buffering capacity [De Frenne et al. 2019; @vonarxSpatiotemporalEffectsForest2012a]. This relationship is reveresed during the night among all forests, however, with pine forests maintaining warmer below canopy temperature than above[@vonarxSpatiotemporalEffectsForest2012a]

@@

[when I think of shade leaves in the understory there are three categories; 1. shade leaves of canopy trees--top of the tree is sun exposed, bottom is shaded. 2. seedlings/saplings of canopy trees that will eventually, if they survive, reach the canopy. 3 understory plants that complete their lifecycle in the shaded understory. These groups are likely to have different traits, and if we expect 'shade' plants to have a different sensitivity to climate change than sun leaves, the consequences of climate change for demography and species community composition of the forest will depend strongly on which category of shade plants we're talking about. I think that this distinction should be addressed throughout the manuscript, so that it is always clear that the reader knows whether the cited references are about inter- or intra-specific variation, and that we address the different trait axes (Marielle made a comment on this earlier as well, that within species patterns may not follow leaf economics spectrum expectations). Ideally we would address the different categories, but if lack of information forces us to focus on just within species or just across species, this should still be made clear. photosynthesis, and canopy-level carbon uptake. They don't have to be treated separately, but it should always be clear which is being referred to. And in the consequences for scaling section this is of course also relevant.

in line with Marielle's comment below; it will be important to distinguish between within plant variation vs across species variation. Understory species may have conservative  growth strategies because they are resource (light) limited. Spending their lifecycle in the shade means they have different pressures than plants that get the bulk of their carbon from sun-exposed leaves, but can maintain a few layers of shade leaves as well. Their traits will be different from shade leaves of canopy trees as well. E.g.
They tend to have long-lived leaves with high LMA while shade leaves of canopy trees have large leaves with low LMA. Those differences in traits are likely to affect their thermal properties as well]--Martijn Slot

Leaf traits are shaped by intra canopy light and height, and show intracanopy plasticity with availability of light. Taller trees tend to be exposed to greater irradiance than lower canopy and shorter understory trees shaded by overstory. Higher irradiance has consequential adaptations in leaf morphology with smaller leaf area index in exposed canopies than shaded lower canopy or canopy interior, and well as having deeper leaf lobes in species such as Quercus. rubra. The table above summarizes the variation of different traits based on light regime, however, some of the traits might differ between species along the vertical profile, thus also varying in leaf thermal sensitivity. [Sack et al. 2016]

Additionally, passing light spots in the forest understory often caused by over canopy vegetation and wind speed called sunflecks cause a different type of morphological adaptation in shaded leaves, making them efficient in capturing 50% of light through sunflecks for photosynthesis.

Leaf traits vary along the vertical canopy profile following the pattern of light availability across the gradient [@poorterMetaanalysisPlantResponses2019b]. Leaves in the upper canopy that are exposed to more light differ in their traits compared to shaded leaves exposed to less light [@niinemetsWorldwideAnalysisWithincanopy2015].
Increasing number of studies point towards the importance of incorporating vertical canopy leaf trait variation in modelling for leaf economic spectrum (LES) [Chen et al. 2020], leaf thermoregulation and thermal sensitivity which will provide a more accurate representation of within-canopy gradient response to increasing temperatures.[@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016; @niinemetsWorldwideAnalysisWithincanopy2015, @chenLeafEconomicsSpectrum2020, REFS)

@@

**Many traits vary with height and/or between sun and shade leaves on the same species (Table 1).**
Here, we focus on traits that are important for shaping leaf temperature ($T_{leaf}$) and thermal sensitivity.

@@

*Leaf area / characteristic dimension*:
- Recent paper on influence on leaf size/shape in PCE – certain aspects of leaf shape were not as relevant as expected (Leigh et al. 2017? https://onlinelibrary.wiley.com/doi/full/10.1111/pce.12857)
- Max stomatal conductance - increases with canopy height (Kenzo et al. 2015)
Max transpiration rate - increases with canopy height (Kenzo et al. 2015).
Leaf width and characteristic dimension influence boundary layer resistance which generally is greater in wider understory leaves than upper canopy.
Additionally, increase in leaf area also influences leaf temperature by maintaining greater leaf temperature than leaves with lower leaf area {Wright et al. 2019}
-Boundary layer resistance and thermal sensitivity also depends of leaf width, leaf size and structure. Increase in leaf width increases stomatal resistance-- dependent on stomtal activity and boundary layer resistance [@fauset_differences_2018] such as lobes, leaf elongation and serration, all dissapate heat more efficiently than non-lobed leaves, wide or non serrated leaves. Leaf lobes are deeper in exposed canopy than in shaded understory, as well as leaf elongation and serration are more pronounced in sun exposed canopies[Vogel 1970; Roth-Nebelsick 2001;
Schuepp 1993; Sack et al. 2006].

>>>>

Light conditions, specifically photosynthetically active radiation (PAR) varies along the vertical gradient with canopy height, canopy structure, across species and forests [@koike_leaf_2001] due to cumulative layers of vegetation. Generally, upper canopies are more exposed to irradiance and limit light through shading to canopy interiors, lower canopy layers and understories [@fauset_differences_2018; @niinemetsPackingPhotosyntheticMachinery2009; Parker 1995; @poorterMetaanalysisPlantResponses2019b]
(Fig. `r fig_NEON_vertical`).
In heterogeneous canopies [with high gap fractions], distance from the outer canopy is a better proxy for light environment than height (Parker 1995) and lower canopy light environments are highly variable, ranging from high light gap to deeply shaded understory regions [@chazdonPhotosyntheticLightEnvironments1984; @tymenQuantifyingMicroenvironmentalVariation2017; @smithSeasonalDroughtrelatedChanges2019]. 
 Tropical forests with dense overstory limit light to understories and subsequent layers, where tropical forest floors receive approximately 1-2%  of the light incident to the top of the canopy [@robertsStomatalBoundaryLayerConductances1990a; @chazdonPhotosyntheticLightEnvironments1984; @parkerConsequencesEnvironmentalHeterogeneity2019], whereas mixed-wood boreal and Northern temperate forest floors receive around 4-5% of full sunlight [Bartemucci et al. 2005].
This difference in light regime along a vertical profile is an important driver of plasticity in leaf traits, leaf physiology, adaptation, and thermal sensitivity along the gradient. Leaves along a vertical intra-canopy gradient thus exhibit a variation in leaf trait adapaptions to a range of high light to shaded-ness [@michaletzEnergeticCarbonEconomic2016; @fausetDifferencesLeafThermoregulation2018; @niinemetsResponsesForestTrees2010, @sackHowStrongIntracanopy2006; @doughtyAreTropicalForests2008; Niinemets and Valladares 2006].

@@

Diurnal moderating capacity of under forest microclimate differs between forests with varying canopy structure, crown density, and seasons. Tropical, and temperate broadleaved and non-pine conifer forests maintain lower daytime maximum temperatures than pine and boreal forests. Increase in crown density and leaf area index decreases light and higher maximum temperature permeability in forests, thus maintaining lower maximum daytime temperature. Thus, pine forests have sparse stands resulting in decreased buffering capacity [De Frenne et al. 2019; @vonarxSpatiotemporalEffectsForest2012a]. This relationship is reveresed during the night among, however, with pine forests maintaining warmer below canopy temperature than above[@vonarxSpatiotemporalEffectsForest2012a]

@@

Leaf traits are shaped by intra canopy light and height, and show intra-canopy plasticity in accordance to intensity and availability of light. Taller trees tend to be exposed to greater irradiance than dense lower canopy and shorter understory trees shaded by overstory. Degrees of high irradiance has consequential adaptations in leaf morphology compared to degrees of low irradiance to shaded-ness [@poorterMetaanalysisPlantResponses2019b; Sack et al 2006] within- and across species [Ishii et al. 2018; @niinemetsWorldwideAnalysisWithincanopy2015]. This dichotomy in leaf traits that vary with light intensity generally differ within intra-canopy, however the degree of intra-canopy plasticity also varies across species and geographically [Sack et al. 2006; Ishii et al. 2018].

Adaptions of traits to ranges of sun-exposure or shaded-ness also vary within a forest strata. 
Understory species may have conservative  growth strategies because they are resource (light) limited. Spending their lifecycle in the shade means they have different pressures than plants that get the bulk of their carbon from sun-exposed leaves, but can maintain a few layers of shade leaves as well. Their traits will be different from shade leaves of canopy trees as well. E.g.
They tend to have long-lived leaves with high LMA while shade leaves of canopy trees have large leaves with low LMA. Those differences in traits are likely to affect their thermal properties as well.
Additionally, shade leaves in the understory are of three categories; 1. shade leaves of canopy trees--top of the tree is sun exposed, bottom is shaded. 2. seedlings/saplings of canopy trees that will eventually, if they survive, reach the canopy. 3 understory plants that complete their lifecycle in the shaded understory. These groups are likely to have different traits, and if we expect 'shade' plants to have a different sensitivity to climate change than sun leaves, the consequences of climate change for demography and species community composition of the forest will depend strongly on the category of shade plants being considered.[refs]

Passing light in the forest understory often caused by over-canopy vegetation and wind speed called sunflecks cause a different type of morphological adaptation in shaded leaves that make them efficient in capturing >50% of light through sunflecks for photosynthesis [@chazdonPhotosyntheticLightEnvironments1984]

[I think that this distinction should be addressed throughout the manuscript, so that it is always clear that the reader knows whether the cited references are about inter- or intra-specific variation, and that we address the different trait axes (Marielle made a comment on this earlier as well, that within species patterns may not follow leaf economics spectrum expectations). Ideally we would address the different categories, but if lack of information forces us to focus on just within species or just across species, this should still be made clear. photosynthesis, and canopy-level carbon uptake. They don't have to be treated separately, but it should always be clear which is being referred to. And in the consequences for scaling section this is of course also relevant.]

Increasing number of studies point towards the importance of incorporating vertical canopy leaf trait variation in modelling for leaf economic spectrum (LES) [Chen et al. 2020], leaf thermoregulation and thermal sensitivity which will provide a more accurate representation of within-canopy gradient response to increasing temperatures.[@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016; @niinemetsWorldwideAnalysisWithincanopy2015, @chenLeafEconomicsSpectrum2020, REFS)

@@

**Many traits vary with height and/or between sun and shade leaves on the same species (Table 1).**
Here, we focus on intra-canopy traits that are important for shaping leaf temperature ($T_{leaf}$) and thermal sensitivity.

@@

*Leaf area / characteristic dimension*:
- Recent paper on influence on leaf size/shape in PCE – certain aspects of leaf shape were not as relevant as expected (Leigh et al. 2017? https://onlinelibrary.wiley.com/doi/full/10.1111/pce.12857)
- Max stomatal conductance - increases with canopy height (Kenzo et al. 2015)
Max transpiration rate - increases with canopy height (Kenzo et al. 2015).
Leaf width and characteristic dimension influence boundary layer resistance which generally is greater in wider understory leaves than upper canopy.
Leaf temperature increases with leaf area {Wright et al. 2019}
-Boundary layer resistance and thermal sensitivity also depends of leaf width, leaf size and structure. Increase in leaf width increases stomatal resistance-- dependent on stomtal activity and boundary layer resistance [@fauset_differences_2018]. Leaf characteristics such as lobes, leaf elongation and serration, all facilitate in dissapating heat more efficiently. Leaf lobes are deeper in exposed canopy than in shaded understory, as well as leaf elongation and serration are more pronounced in sun exposed canopies[Vogel 1970; Roth-Nebelsick 2001;
Schuepp 1993; Sack et al. 2006].
